\bf Model Visualization Techniques for a Social Network Model

### 2017-06-06 paper.Rnw
add some edits. some comments i don't quite understand

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators. Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
}

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\alpha$. For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$. \hh{XXX that means that the function is essentially just a rate parameter, right?}
Other modelling scenarios allow this rate to be more flexible,  e.g.\  be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
\hh{How is the rate parameter different from a rate parameter?}
The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$. We assume that the actors $i$ are conditionally independent given their current ties, $x_{i1}, \dots, x_{in}$. This assumption leads to the rate for the whole network, $n\alpha$. In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$

@@

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time. The objective function is then written as \hh{the sum of the individual changes XXX write out the math in words XXX} $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates. The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.

The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$. \hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p. 371]{snijders01}. \hh{'There are many' another vague phrase} There are many possible parameters $\beta_i$ to add to the model. \hh{'They' ... be specific.} They can be split up into two groups: first, the structural effects, which only depend on the structure of the network. \st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$. The second set of effects are the actor-level or covariate effects.} \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} These effects also depend on the structure of the network. They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can be included \hh{currently} in the objective function, see \citet{RSiena}.

>>>>

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically with traditional methods. Likelihood functions turn into very complex objects due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators. Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
}

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$. For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$. %\hh{XXX that means that the function is essentially just a rate parameter, right?}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
%\hh{How is the rate parameter different from a rate parameter?}
In our model with the simple rate function, the rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period from $t_{m}$ to $t_{m+1}$. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau_i(m)$ be the wait time until actor $i$ makes its next change.For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

\begin{equation} \label{eq:c1}
 \tau_i(m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}

 The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$. The waiting time for the whole network to change, $\tau(m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as

\begin{equation} \label{eq:c2}
 \tau(m) \sim Exp(n\alpha_m)
\end{equation}

} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs. %In order to achieve the memorylessness property of a Markov process,

@@

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional independence assumptions in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, since only one tie from one node is changing at a time. The objective function is then written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates. The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.

The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$. %\hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation only depend on the structure of the network, like the outdegree and reciprocity parameters mentioned above. %\hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} 
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, and are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.
